Identifying Microlenses in Large, Non-Uniformly Sampled Surveys: The Case of the Palomar Transient Factory

### 2012-08-10 text/ptf_microlensing_paper.tex

%\documentclass{emulateapj}
\documentclass[12pt,preprint]{aastex}
\newcounter{address}
\setcounter{address}{1}
\usepackage{lscape, longtable}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{subfigure}
\usepackage{natbib}
\newcommand{\Msun}{\ifmmode {M_{\odot}}\else${M_{\odot}}$\fi}
\newcommand{\Rsun}{\ifmmode {R_{\odot}}\else${R_{\odot}}$\fi}
\newcommand{\lapprox }{{\lower0.8ex\hbox{$\buildrel <\over\sim$}}}
\newcommand{\gapprox }{{\lower0.8ex\hbox{$\buildrel >\over\sim$}}}
\def\amin{\ifmmode^{\prime}\else$^{\prime}$\fi}
\def\asec{\ifmmode^{\prime\prime}\else$^{\prime\prime}$\fi}
\newcommand{\apwsim}{\raisebox{0.2ex}{\scriptsize$\sim$\normalsize}}

@@

The PTF camera is the 12K$\times$8K mosaic camera formerly at the Canada-France-Hawaii Telescope. The camera has 11 working chips, 10$^7$ pixels, and a 7.26 deg$^2$ field-of-view \citep{rahmer2008}. Under median seeing conditions (1.1$\arcsec$), observations in Mould $R$ or Sloan Digital Sky Survey (SDSS) $g$ achieve 2.0$\arcsec$ full-width half-maximum images and reach 5$\sigma$ magnitudes of $R \approx 21.0$ and $g \approx 21.3$ mag in a standard 60~s exposure \citep{nick2010}. As of 2012 May, the PTF footprint included $\sim$9800 (2500) deg$^2$ imaged $>$25 ($>$100) times in $R$ and $\sim$2400 (250) deg$^2$ imaged that often in $g$.

@@

\section{Microlensing event recovery}
Microlensing surveys typically use difference image analysis (cite Lupton?) to initially identify transient events and then analyze the candidate light curves to search for microlensing events (cite MACHO, OGLE). This method is essential for studying crowded fields such as the Galactic bulge, and also limits the number of light curves that must be searched for microlensing events, thus saving on computation time. This [method] also provides real-time detection of events so surveys can utilize networks of telescopes to provide follow-up observations of these transients. Our approach is to instead take the preexisting PTF light curve database over the entire survey footprint and develop fast, robust search algorithms to pick out microlensing event candidates from this massive data archive.

@@

>>>>

%\documentclass{emulateapj}
\documentclass[12pt,preprint]{aastex}
\newcounter{address}
\setcounter{address}{1}
\usepackage{lscape, longtable}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{subfigure}
\usepackage{natbib}
\newcommand{\Msun}{\ifmmode {M_{\odot}}\else${M_{\odot}}$\fi}
\newcommand{\Rsun}{\ifmmode {R_{\odot}}\else${R_{\odot}}$\fi}
\newcommand{\lapprox }{{\lower0.8ex\hbox{$\buildrel <\over\sim$}}}
\newcommand{\gapprox }{{\lower0.8ex\hbox{$\buildrel >\over\sim$}}}
\def\amin{\ifmmode^{\prime}\else$^{\prime}$\fi}
\def\asec{\ifmmode^{\prime\prime}\else$^{\prime\prime}$\fi}

\newcommand{\apwsim}{\raisebox{0.2ex}{\scriptsize$\sim$\normalsize}} 
\newcommand{\inlinecode}{\texttt}

@@

The PTF camera is the 12K$\times$8K mosaic camera formerly at the Canada-France-Hawaii Telescope. The camera has 11 working chips, 10$^7$ pixels, and a 7.26 deg$^2$ field-of-view \citep{rahmer2008}. Under median seeing conditions (1.1$\arcsec$), observations in Mould $R$ or Sloan Digital Sky Survey (SDSS) $g$ achieve 2.0$\arcsec$ full-width half-maximum images and reach 5$\sigma$ magnitudes of $R \approx 21.0$ and $g \approx 21.3$ mag in a standard 60~s exposure \citep{nick2010}. As of August 9, 2012, the PTF footprint included $\apwsim$10,100 (2600) deg$^2$ imaged $>$25 ($>$100) times in $R$ and $\apwsim$2200 (170) deg$^2$ imaged that often in $g$.

@@

\section{Microlensing event recovery}
Microlensing surveys typically use difference image analysis (DIA;cite Lupton?) to identify transient events and then analyze light curves of these transient sources to search for microlensing events (cite MACHO, OGLE). This method is essential for studying crowded fields such as the Galactic bulge where photometry is difficult (cite or figure?), but also limits the number of light curves that must be searched for microlensing events, thus saving on computation time. Pre-selection with DIA also enables fast, real-time detection of events so dedicated microlensing surveys can obtain high-cadence follow-up observations of transient sources. Our approach is to instead take the preexisting PTF light curve database over the entire survey footprint and develop fast, robust search algorithms to pick out microlensing event candidates from this massive data archive.

The average number of light curves with $>$25 ``good'' observations is $\apwsim 50,000$ per PTF field, with $\apwsim64,000,000$  ``good'' light curves in total. By our estimates, utilizing all cores on an 8-core 2.0 GHz compute, it would take [TODO: estimate this number] to fit even the simplest 5-parameter microlensing event model to each of these light curves using \inlinecode{ANSI-C} wrapped in \inlinecode{Python}. Previous studies have had the advantage of pre-selecting candidates via difference image analysis, but have utilized a variety of selection techniques for flagging candidate microlensing light curves after this step. However, most of these methods still rely on fitting a point-source, point-lens microlensing model to each light curve and performing some selection based on the $\chi^2$ value of this fit (cite Sumi et al. 2006, MACHO work, ?). More recent work has begun to apply machine learning algorithms to the general light curve classification problem (cite Joey Richards), however this is still a work in progress. We set out to find a fast, robust selection method that [TODO:].

@@

% Talk about hypothetical selection criteria, and the things we have to consider for detection efficiency (possible selection criteria, optimizing selection method, selecting candidates)
